---
title: "p8105_hw2_gp2770"
author: "Gokul Pareek"
date: "2024-10-03"
output: github_document
---

I'm an R Markdown document!

# Problem 1

Reading and Cleaning the Data

```{r}
library (tidyverse)

nyc_transit_data <- read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",col_types= cols(Route8 = "c", Route9 = "c",Route10 = "c",Route11 = "c")) %>%   janitor::clean_names() %>%
  select(line, station_name, station_latitude, station_longitude,route1:route11, entry, vending, entrance_type, ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

nyc_transit_data

```

The resulting dataset contains information about each entrance and exit for each subway station in NYC, including details such as the subway line, station name, geographic coordinates, routes served, entrance type, and ADA compliance status. Till now, I have standardized the variable names by converting them to lower snake case, selected and retained the relevant variables, and recoded the entry variable from a character class with "YES" and "NO" values to a logical class with "TRUE" and "FALSE" values. The dataset currently has 1,868 rows and 19 columns. However, the data is not yet tidy, as the multiple columns for route numbers (routes 1 through 11) represent values themselves, rather than being captured as a single variable. 

Finding the number of distinct stations

```{r}

nrow(distinct(nyc_transit_data, line, station_name))

```

There are 465 distinct stations.

ADA compliant stations

```{r}

nyc_transit_data %>% 
  filter(ada == TRUE) %>% 
  distinct(line, station_name) %>% 
  nrow()

```

Out of the 465 distinct stations, 84 stations are ADA compliant. 

Proportion of station entrances/exits without vending that allow entrance

```{r}

without_vending <- nyc_transit_data %>% 
                    filter(vending == "NO") %>%
                    filter(entry == TRUE) %>%  
                    nrow()

with_vending <- nyc_transit_data %>% 
                  filter(vending == "NO") %>%
                  nrow()

without_vending / with_vending

```

0.377 is the proportion of station entrances/exits without vending that allow entrances. 

Reformatting the data to make route number and route name distinct variables

```{r}

nyc_transit_reformat_data <- nyc_transit_data %>%
  pivot_longer(route1:route11, 
               names_to = "route_number", 
               values_to = "route_name")

nyc_transit_reformat_data

```

Distinct stations that serve the A train

```{r}

nyc_transit_reformat_data %>% 
  filter(route_name == "A") %>% 
  distinct(line, station_name) %>% 
  nrow()

```

60 distinct stations serve the A train.

A train serving stations that are ADA compliant

```{r}

nyc_transit_reformat_data %>% 
  filter(ada == TRUE) %>%  
  filter(route_name == "A") %>% 
  distinct(line, station_name) %>% 
  nrow()
```

Of the 60 stations that serve the A train, 17 stations are ADA compliant.

# Problem 2

Reading and Cleaning the Mr. Trash Wheel sheet

```{r}

library(readxl)

mr_trash_wheel_data <- read_excel("./data/202409 Trash Wheel Collection Data.xlsx",                           sheet = "Mr. Trash Wheel",range =  cell_cols("A:N"), 
                        col_names = TRUE, skip = 1) %>%
  janitor::clean_names() %>%
  rename(
    collection_month = month,
    collection_year = year,
    plastic_bottles_collected = plastic_bottles,
    polystyrene_collected = polystyrene,
    cigarette_butts_collected = cigarette_butts,
    sports_balls_collected = sports_balls,
    glass_bottles_collected = glass_bottles,
    wrappers_collected = wrappers,
    plastic_bags_collected = plastic_bags
    )%>%
  filter(!is.na(dumpster)) %>%
  mutate(
    sports_balls_collected = as.integer(round(sports_balls_collected)),
    trash_wheel = "Mr. Trash Wheel"
  )

```

Reading and Cleaning the Professor Trash Wheel

```{r}

professor_trash_wheel_data <-  read_excel("./data/202409 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = cell_cols("A:M"), col_names = TRUE, skip = 1) %>%
  janitor::clean_names() %>%
  rename(
    collection_month = month,
    collection_year = year,
    plastic_bottles_collected = plastic_bottles,
    polystyrene_collected = polystyrene,
    cigarette_butts_collected = cigarette_butts,
    glass_bottles_collected = glass_bottles,
    wrappers_collected = wrappers,
    plastic_bags_collected = plastic_bags
    )%>%
  filter(!is.na(dumpster)) %>%
  mutate(trash_wheel = "Professor Trash Wheel"
  )

```

Reading and Cleaning the Gwynnda Trash Wheel

```{r}

gwynnda_trash_wheel_data <-  read_excel("./data/202409 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = cell_cols("A:L"), col_names = TRUE, skip = 1) %>%
  janitor::clean_names() %>%
  rename(
    collection_month = month,
    collection_year = year,
    plastic_bottles_collected = plastic_bottles,
    polystyrene_collected = polystyrene,
    cigarette_butts_collected = cigarette_butts,
    wrappers_collected = wrappers,
    plastic_bags_collected = plastic_bags
    )%>%
  filter(!is.na(dumpster)) %>%
  mutate(trash_wheel = "Gwynnda Trash Wheel"
  )

```

Combining the three datasets

```{r}

mr_trash_wheel_data <- mr_trash_wheel_data %>%
  mutate(collection_year = as.character(collection_year))

professor_trash_wheel_data <- professor_trash_wheel_data %>%
  mutate(collection_year = as.character(collection_year))

gwynnda_trash_wheel_data <- gwynnda_trash_wheel_data %>%
  mutate(collection_year = as.character(collection_year))

combined_trash_wheel_data <- bind_rows(mr_trash_wheel_data, 
                                       professor_trash_wheel_data, 
                                       gwynnda_trash_wheel_data)

combined_trash_wheel_data

```

The combined Trash Wheel dataset contains a total of `r nrow(combinedtrash_wheel_data)` observations, consolidating information from three different Trash Wheels: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Key variables include `collection_month`, `collection_year`, `weight_tons` which is the weight of trash collected by each dumpster, and waste types like `cigarette_butts_collected` and `plastic_bags_collected`.

Total weight of trash collected by Professor Trash Wheel

```{r}

total_professor_trash_wheel_weight <- combined_trash_wheel_data %>%
  filter(trash_wheel == "Professor Trash Wheel") %>%
  pull(weight_tons) %>%
  sum(na.rm = TRUE)

total_professor_trash_wheel_weight

```

The total weight of trash collected by Professor Trash Wheel was 246.74 tons.

Total number of cigarette butts collected by Gwynnda in June of 2022

```{r}

total_cigarette_butts_gwynnda_june_2022 <- combined_trash_wheel_data %>%
  filter(trash_wheel == "Gwynnda Trash Wheel" & collection_month == "June" & collection_year == 2022) %>%
  pull(cigarette_butts_collected) %>%
  sum(na.rm = TRUE)

total_cigarette_butts_gwynnda_june_2022

```

The total number of cigarette butts collected by Gwynnda in June of 2022 was 18120.

